`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages

### 2016-04-29 response-to-reviewers.tex

\respquote{We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
species. 
We used the same subset of the morphological data as in \cite{gavryushkina2015bayesian} but we did not use the molecular sequence data from the living species. 
The morphological data matrix we used contains 36 fossil species, 19 extant species and 202 characters (ranging from binary to k=7). 
The majority of these characters (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
See \cite{gavryushkina2015bayesian} for further details of data selection.}

Maybe add a small "phylogentics analysis" section, between p.4 and p.5:

Here you can describe the analysis steps that are common to both datasets. Please provide brief descriptions of models M1 and M8. I note that in Lewis' Mk model, the k stands for the dimension of the character substitution matrix, so M2 means a 2x2 substitution matrix (binary characters). This does not seem to be the case here. The authors could either change their notation slightly (suggestion: Mk-1, Mk-8), or explicitly state that their nomenclature does not follow Lewis.

\response{We added a section called Phylogenetic analyses, explaining the stepwise analysis of fossil age as well as the morphological models M1 and M8}

%\respquote{Following \cite{gavryushkina2015bayesian}, we apply the Lewis Mk model \cite{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states. As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered. 
%We applied two phylogenetic models to the penguin data set: Mk-1 and Mk-8. Mk-1 assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites. Model Mk-8 partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.}

@@

Then, in stepwise analysis, each fossil age is in turn represented not by the stratigraphic uniform distribution, but by a prior density derived from the fossilised birth-death process.

\response{No, each fossil is in turn analyzed using the prior density derived from the fossilised birth-death process without conditioning on its stratigraphic bounds, while the remaining fossils are conditioned on their stratigraphic bounds. This is detailed in the new ``Phylogenetic analyses'' section. }

 Give some details please.

 \response{We have added the opening two paragraphs in the new ``Phylogenetic analyses'' section that give the details.}

@@

where $p(tip\_ages)$ is the prior density on fossil ages using uniform distributions, $p(node\_ages)$ is the prior on internal node ages given the BD process, and $L(D|ages,tree)$ is the likelihood of the morphological data D, given the ages and the topology calculated using Lewis (2001) formula. However, it seems from the text in p.5 that another likelihood term may be involved. For the stepwise deletion procedure, I am guessing the priors are something like $p(tip\_ages\hat{}) p(node\_ages, tip\_age*)$, where now $tip\_ages\hat{}$ includes all the fossil ages except the one under analysis, which has been moved into the other BD-based prior term $(tip\_age*)$. Please provide general forms for the posteriors and make the connection with equation (1).

\response{We have now provided a general form of the posterior we are sampling under the new section on ``Phylogenetic analyses''. }

Give some more info on how the prior on the root age is set. Is Tmax the same as the root age? If it is, please state explicitly.

\response{$T$ is not the age of the root, it is the origin of the fossilized birth-death process, and is older than the root. It can be interpreted as the stem age of the clade.}

@@

Please give the prior on the morphological substitution rate, and for relaxed-clock models, the prior on the rate variation parameter (I am assuming you are using log-normal clock, so prior on log-variance).

\response{We have added the prior information for the morphological substitution rate and log standard deviation of the relaxed clock.}

p.8 paragraph starting on line 43. Do you mean that with the particular implementation of marginal likelihood calculation in BEAST you cannot calculate the BF? Or is this a general problem with any implementation of stepping-stones and this model?

\response{TODO - thinking about this: why can we not use stepping stone? We could have calulated the marginal likelihood given the focal fossil is in the paleontological range, and the marginal likelihood given the fossil is outside the range, no?}

I am also confused, do sampling times refer to the ages of fossil tips? Aren't these unknown parameters being estimated in the MCMC? Then they would appear as terms in the prior and the likelihood. (?)

@@

As I pointed out at the beginning of this review, many papers report ancient estimates of node ages using morphological data. How do the posterior age of the root for canids and penguins obtained in this paper compare with estimates from traditional methods? Please mention briefly in discussion. For example, some of the previous estimates could be added to the table I suggested.

\response{The age of extant penguins in the subject of \cite{gavryushkina2015bayesian}, so we don't discuss it here, however, the estimate obtained is much younger than previous published estimates using node-dating, contrary to the reviewer's suggestion that morphological data tends to produce ancient estimates.}

I think the insistence by the authors to use the term "morhpological clock" will cause controversy. Please provide some evidence that departure of the clock is small (for example, posterior estimates of the rate variation parameter are, say, close to zero). Otherwise I suggest removing the term.

\response{The estimated log sd of the evolutionary rate is 0.69 in the penguin Mk-8 analysis (see Table 3), which is not much different to estimate of the same parameter for relaxed molecular clocks in, for example, influenza.}

### 2016-04-29 fossilDating-revision.tex
Tanja revision start

%TSrev
Given the morphological data $D$ and a stratigraphic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratigraphic age range. The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the model for morphological character evolution are summarized in $\theta$. More formally, we sample from,
$$P[\mathcal{T},\eta, \theta | D, a] = P[D|\mathcal{T},\theta] P[\mathcal{T},a|\eta] P[\eta] P[\theta]/P[D,a],$$
where $ P[\mathcal{T},a|\eta] =  P[\mathcal{T}|\eta] $ if each fossil age is within its stratigraphic age range specified in $a$, and $ P[\mathcal{T},a|\eta] =0$ else.
When we replaced the focal  palaeontological age constraints of fossil $i$  by 
the fossilized birth-death process prior, we simply set $l_i=0$ and $u_i=T$.

>>>>

%TSrev
Given the morphological data $D$ and a stratigraphic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratigraphic age range (i.e. the phylogenetic estimate of fossil age). The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the model for morphological character evolution are summarized in $\theta$. More formally, we sample from,
$$P[\mathcal{T},\eta, \theta | D, a] = P[D|\mathcal{T},\theta] P[\mathcal{T},a|\eta] P[\eta] P[\theta]/P[D,a],$$
where $ P[\mathcal{T},a|\eta] =  P[\mathcal{T}|\eta] $ if each fossil age is within its stratigraphic age range specified in $a$, and $ P[\mathcal{T},a|\eta] =0$ else.
When we replaced the focal  palaeontological age constraints of fossil $i$  by 
the fossilized birth-death process prior, we simply set $l_i=0$ and $u_i=T$.
